---
title: "Cohort 11 — B-cell abundance as predictor of ICB response in HNSCC "
author: "Marie Wyrsch"
date: today

toc: true
abstract: "Reproducing and extending: Chang et al. (2025), Annals of Oncology,  DOI: https://doi.org/10.1016/j.annonc.2024.09.014"

format:
  html:
    embed-resources: true #this creates a clean HTML file
    number-sections: true #sections are numbered
    code-fold: show #hide or show code 

editor: visual
editor_options:
  chunk_output_type: console

execute:
  warning: false
  message: false
---

## Introduction

For our reproducibility study we used the following variables:

- Predictor: B_cell_abundance (B cells as % of PBMCs, flow cytometry)

- Outcome: best_response (CR/PR = responder; SD/PD = non-responder)

- Confounder: tnm_stage (IV vs non-IV; model 2 only)

- Method: Firth penalized logistic regression (due to complete separation)

```{r}
#| label: load-and-prepare-data
#| output: false
#| code-fold: false



#loading of libraries ----------------------------------------------------------------
library(readxl)
library(brglm2)   # Firth penalized logistic regression
library(pROC)     # ROC curves and AUC
   


#import data -------------------------------------------------------------------------
library(readxl)

path <- "/Users/mariewyrsch/Documents/UZH/BME321/mini_project/BME321-Mini-Project/Data_of_in-house_cohorts.xlsx"

print(file.exists(path))
print(readxl::excel_sheets(path))

df <- readxl::read_excel(
  path,
  sheet = "Cohort11",
  skip = 1
)

head(df)
colnames(df)
```

The relevant columns (adjust names if your read produces different strings):

- Flow B Cells Count: raw B-cell count from flow cytometry

- Flow Lymphocytes Count: lymphocyte count

- Flow Monocytes Count: monocyte count

- Best response: RECIST response category

- TNM stage: clinical TNM stage

## Compute B-Cell abundance

For the computation the definition in the paper was followed

Definition (per paper): B cells / (lymphocytes + monocytes) \* 100

```{r}
#| label: computing-b-cell-abundandce
#| output: true


df$B_cell_abundance <- df[["Flow B Cells Count"]] /
  (df[["Flow Lymphocytes Count"]] + df[["Flow Monocytes Count"]]) * 100

summary(df$B_cell_abundance)
```

```{r}
#| label: computing-t-cell-abundance
#| output: true

df$t_cell_abundance <- (df[["Flow T CD4+ Count"]] + df[["Flow T CD8+ Count"]]) /
  df[["Flow Lymphocytes Count"]]
summary(df$t_cell_abundance)
```

## Binarize Outcome

The next step was also done according to the paper. The outcome variables is binarized as followed: CR or PR = responder (1), SD or PD = non-responder (0)

The expected outcome was: 8 responders, 52 non-responders

And after binarizing it was checked whether there were no responders below the threshold of B-Cell abundance below 5.5%.

```{r}
#| lable: binarizing-outcome
#| output: true
#| code-fold: true

df$best_response <- ifelse(
  df[["Best response"]] == "CR" | df[["Best response"]] == "PR",
  1,
  0
)
table(df$best_response)
# Expected: 8 responders, 52 non-responders

# Cross-check with 5.5% threshold from paper (should show 0 responders below)
table(
  "B >= 5.5%" = df$B_cell_abundance >= 5.5,
  "Responder" = df$best_response
)
```

## Collapse TNM stage

```{r}
#| label: TNM-stage-distribution
#| output: true
#| echo: false

tnm_counts <- df %>%
  count(`TNM stage`, name = "n")

tnm_counts$`TNM stage` <- factor(
  tnm_counts$`TNM stage`,
  levels = c("I", "II", "IIA", "III", "IVA", "IVB", "IVC", "IV")
)

ggplot(tnm_counts, aes(x = n, y = `TNM stage`)) +
  geom_col(width = 0.4, fill = "#0F7C7B") +
  geom_text(aes(label = n), hjust = -0.5, size = 5, color = "#2E3A4D") +
  scale_x_continuous(
    limits = c(0, max(tnm_counts$n) + 2),
    breaks = seq(0, max(tnm_counts$n) + 2, by = 5),
    expand = c(0, 0)
  ) +
  labs(
    title = "TNM Stage Distribution",
    x = "Number of patients",
    y = NULL
  ) +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", size = 16, color = "#1F2A44"),
    axis.text = element_text(color = "#2E3A4D", size = 14),
    axis.title.x = element_text(color = "black", size = 13),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "#DDE3EA"),
    axis.line = element_line(color = "gray50"),
    axis.ticks = element_line(color = "gray50"),
    panel.background = element_rect(fill = "#F3F7F8", color = NA),
    plot.background = element_rect(fill = "white", color = NA)
  )
```

TNM stages were collapsed/grouped as follows:

- stage IV (any subtype) = 1

- non-IV = 0

Reason for this decision was the fact that 57 out of 60 patients are stage III-IV and with that too few per level for a multi-level factor with only 8 responders due to risk of overfitting.

```{r}
#| label: collapse-tnm-stage
#| output: true

df$tnm_stage_IV <- as.integer(grepl("^IV", df[["TNM stage"]]))

df$tnm_stage_IV <- ifelse(
  grepl("^IV", df[["TNM stage"]]),
  1,
  0
)

table(df[["TNM stage"]], df$tnm_stage_IV, useNA = "ifany")

# 3 patients have missing TNM stage — excluded from model 2

```

## Separation Check

Standard logistic regression fails when a predictor perfectly separates the outcome. Here all 8 responders have B_cell_abundance \> 5.5%.

```{r}
#| lable: fig-separation-check
#| fig-cap: "B-Cell abundance by response group"
#| output: true

par(mar = c(5, 5, 4, 2))

boxplot(
  B_cell_abundance ~ best_response,
  data   = df,
  names  = c("Non-responders", "Responders"),
  ylab   = "B-cell abundance (% of PBMCs)",
  main   = "B-cell abundance by response group",
  col    = c("lightcoral", "lightblue")
)

stripchart(
  B_cell_abundance ~ best_response,
  data   = df,
  method = "jitter", jitter = 0.08,
  vertical = TRUE, add = TRUE,
  pch = 16, col = c("lightcoral", "lightblue"), cex = 0.8
)

abline(h = 5.5, lty = 2, col = "gray50")
```

```{r}
#| label: distributional-statistics_response
#| output: true
library(dplyr)

df %>%
  group_by(best_response) %>%
  summarise(
    n = n(),
    mean = mean(B_cell_abundance, na.rm = TRUE),
    sd = sd(B_cell_abundance, na.rm = TRUE),
    median = median(B_cell_abundance, na.rm = TRUE),
    IQR = IQR(B_cell_abundance, na.rm = TRUE),
    min = min(B_cell_abundance, na.rm = TRUE),
    max = max(B_cell_abundance, na.rm = TRUE)
  )

df$below_5_5 <- df$B_cell_abundance < 5.5

df %>%
  group_by(best_response) %>%
  summarise(
    n_total = n(),
    n_below = sum(below_5_5, na.rm = TRUE),
    pct_below = 100 * mean(below_5_5, na.rm = TRUE)
  )



```

```{r}
#| label: fig-t-cell-separation
#| fig-cap: "T-cell abundance by response group"

boxplot(
  t_cell_abundance ~ best_response,
  data   = df,
  names  = c("Non-responders", "Responders"),
  ylab   = "T-cell abundance abundance (% of PBMCs)",
  main   = "T-cell abundance by response group",
  col    = c("lightcoral", "lightblue")
)

stripchart(
  t_cell_abundance ~ best_response,
  data   = df,
  method = "jitter", jitter = 0.08,
  vertical = TRUE, add = TRUE,
  pch = 16, col = c("lightcoral", "lightblue"), cex = 0.8
)
```

## Model 1 - Univariate Firth logistic regression (reproduce paper)

If a predictor perfectly predicts the outcome:

- Example: all patients with a biomarker = 1 have the event, none without it do

```{r}
#| label: model-1
#| code-fold: false
model1 <- glm(
  best_response ~ B_cell_abundance,
  data   = df,
  family = binomial(link = "logit"),
  method = "brglmFit"             # Firth penalization via brglm2
)

summary(model1)
exp(coef(model1))                 # odds ratios
exp(confint(model1))              # 95% CI on odds ratios

# ROC curve and AUC
roc1 <- roc(
  response  = df$best_response,
  predictor = fitted(model1),
  ci        = TRUE,
  quiet     = TRUE
)

cat("Model 1 — Univariate AUC:", round(auc(roc1), 4), "\n")
cat("95% CI:", round(ci(roc1)[1], 4), "–", round(ci(roc1)[3], 4), "\n")
# Target from paper: AUC ≈ 0.73
```

```{r}
#| label: model_t_cell
t_model <- glm(
  best_response ~ t_cell_abundance,
  data   = df,
  family = binomial(link = "logit"),
  method = "brglmFit"             # Firth penalization via brglm2
)

summary(t_model)
exp(coef(t_model))                 # odds ratios
exp(confint(t_model))              # 95% CI on odds ratios

# ROC curve and AUC
roct <- roc(
  response  = df$best_response,
  predictor = fitted(t_model),
  ci        = TRUE,
  quiet     = TRUE
)

cat("T-Cell Model — Univariate AUC:", round(auc(roct), 4), "\n")
cat("95% CI:", round(ci(roct)[1], 4), "–", round(ci(roct)[3], 4), "\n")
# Target from paper: AUC ≈ 0.73
```

## Model 2 - adjusted for TNM

```{r}
#| label: model-2
#| code-fold: false

df_complete <- df[!is.na(df[["TNM stage"]]), ]   # n = 57 (3 missing TNM)
cat("n for adjusted model:", nrow(df_complete), "\n")

model2 <- glm(
  best_response ~ B_cell_abundance + tnm_stage_IV,
  data   = df_complete,
  family = binomial(link = "logit"),
  method = "brglmFit"
)

summary(model2)
exp(coef(model2))
exp(confint(model2))

roc2 <- roc(
  response  = df_complete$best_response,
  predictor = fitted(model2),
  ci        = TRUE,
  quiet     = TRUE
)

cat("Model 2 — Adjusted AUC:", round(auc(roc2), 4), "\n")
cat("95% CI:", round(ci(roc2)[1], 4), "–", round(ci(roc2)[3], 4), "\n")

```

## Compare AUCs

For the comparison a DeLong's test was used as in the paper.

Note that the models use different n (60 vs 57) because of that the comparison was made on shared complete cases.

```{r}
#| label: compare-AUC
#| output: true


model1b <- glm(
  best_response ~ B_cell_abundance,
  data   = df_complete,
  family = binomial(link = "logit"),
  method = "brglmFit"
)
roc1b <- roc(df_complete$best_response, fitted(model1b), quiet = TRUE)

roc_comparison <- roc.test(roc1b, roc2, method = "delong")
print(roc_comparison)
```

```{r}
#| label: fig-auc-compare
#| output: true
#| echo: false


auc_df <- tibble(
  model = c("Our + TNM adj.", "Our univariate", "Paper (Fig 4D)"),
  auc = c(
    auc(roc2),
    auc(roc1),
    0.73  # paper value 
  )
)

ggplot(auc_df, aes(x = auc, y = model, fill = model)) +
  geom_col(width = 0.4) +
  geom_text(aes(label = sprintf("%.3f", auc)), hjust = -0.2) +
  scale_fill_manual(values = c(
    "Our + TNM adj." = "lightcoral",
    "Our univariate" = "lightblue",
    "Paper (Fig 4D)" = "#5D6D7E"
  )) +
  coord_cartesian(xlim = c(0.5, 0.85)) +
  theme_minimal() +
  theme(legend.position = "none")

```

## Plot ROC curves

```{r}
#| label: fig-plot-roc-curves
#| output: true

par(mar = c(5, 4, 3, 2))

plot(
  roc1,
  col  = "lightblue",
  lwd  = 2,
  xlab = "1 - Specificity",
  ylab = "Sensitivity"
)

plot(
  roc2,
  col  = "lightcoral",
  lwd  = 2,
  lty  = 2,
  add  = TRUE
)

plot(
  roct,
  col  = "lightgreen",
  lwd  = 2,
  lty = 2,
  add = TRUE
)

abline(a = 0, b = 1, lty = 3, col = "gray60")

legend(
  "bottomright",
  legend = c(
    paste0("Univariate  (AUC = ", round(auc(roc1),  3), ")"),
    paste0("+ TNM stage (AUC = ", round(auc(roc2),  3), ")"),
    paste0("Univariate T-Cell (AUC =", round(auc(roct), 3), ")"),
    "Random (AUC = 0.500)"
  ),
  col    = c("lightblue", "lightcoral", "lightgreen", "gray60"),
  lty    = c(1, 2, 2, 3),
  lwd    = c(2, 2,2, 1),
  bty    = "n",
  cex    = 0.85
)

```

## Session Info

```{r}
#| label: session-info
#| output: true


sessionInfo()
```
